---
title: "Home team advantage in the NHL"
author: "Samrat Halder, Hariz Johnson, Jake Stamell"
date: "12/6/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Project Overview

### Background and research questions

Home team advantage is a common phenomenon in sports. Intuitively, we expect a team to perform better in their home arena. They are used to that location, have their fans to support them, don’t need to travel to it, etc.

But what does it really mean to have an advantage? Looking below the level of wins and losses, what happens to teams when they are at home and away? Does their style of play change? Are there some places that are harder to play at than others? Are there teams that perform better when away compared to others?

For our project, we explored these questions for ice hockey games in the National Hockey League (NHL). In recent years, the NHL has started tracking where and when events (like shots and goals) happen during the game. This provides a new opportunity to explore the game differently. Not only can we examine high-level game information, but we can also look at play-level details like where events occur on the ice.

This project consists of the following report, which condenses and summarizes our findings. The first section provides a high-level overview of our project, while the second walks through the details of our analysis. We have also developed an interactive shiny app to allow the reader to explore different views of the data. The report primarily focuses on high level views, while the app gives the opportunity to explore more granular views of the data. We encourage the reader to peruse the app after reading through the report.

All code used for the project is availabe on the github page: [TBU].

### What you need to know about hockey to understand this

If you are familiar with the sport, feel free to skip this section. For those unfamiliar with the sport, we provide a brief overview of the relevant information on hockey and the NHL to be able to understand this report. There are penalties…

Ice hockey is a sport where each team has 6 players: 5 skaters and 1 goalie. Teams play 3 periods, each taking 20 minutes. The team with the most goals at the end wins. On average, teams score 2-3 goals per game.

For reference, here are the main types of plays in hockey (these are the ones recorded by the NHL).

Offensive plays

  * Goal: where one team scores
  * Shot: where a team shoots the puck on goal
  * Missed shot: where a team shoots the puck but misses the goal

Defensive 

  * Blocked shot: where a team shoots the puck, but the other team blocks it before it reaches the goal
  * Takeaway: when a team takes the puck away from the other team
  * Hit: when one player hits another (it's allowed and all players wear protective gear!)

Other

  * Penalty: where a foul is called on a team (following a penalty, that team plays with one less player for a short period of time)
  * Giveaway: when a team loses the puck
  * Faceoff: these events start a play and happen in 9 fixed locations on the ice (similar to a tip off in basketball)
  
For more information, check out the wikipedia page for ice hockey: https://en.wikipedia.org/wiki/Ice_hockey.

The NHL is the professional hockey league in the United States. It consists of 31 teams (30 before 2017) that each play 82 games per year. Each team plays half of their games at home and half at other teams’ arenas. (See the wikipedia page for more info: https://en.wikipedia.org/wiki/National_Hockey_League).

### Data collection

We gathered game data for all regular season games for the seasons in 2014-2018. This data consists of general information about the games (6232 total games) such as which teams played, who won, how many shots, where it was, etc. Additionally, we collected play-level data of each major event in the game. This includes the location, team, players, and more of each play in each game for events like goals, shots, penalties, etc. This is the primary 
data set for our project, consisting of (~1.6M events).

We collected our data directly from the NHL.com API, using the httr package to make requests and the jsonlite package to parse the raw json data. While there is no formal documentation for this API, there is an online community that has documented some of the available endpoints (https://gitlab.com/dword4/nhlapi). From there, we sifted through the information returned by these calls to find the information we were looking for.


To recreate this dataset, visit our github for the project and use script "2019-12-06_nhl-api-script.R". The data itself is also directly available in the RData file "2019-12-07_nhl-cleaned-data.RData".

### High level findings

In the NHL, ~55% of games are won by the home team. While the home team has a clear advantage, the degree of it varies by team. (Even if a really good team is playing away, they still are likely to be the better team.) Digging in to the play-level data, we see that home team advantage exists across almost all types of plays. Not only do teams shoot and score (and thus win) more when they are the home team, they also are penalized less. Perhaps fans yelling at the referee does have some effect!

Hardest places to play: Anaheim and Washington

Easiest places to play: Ottowa and Edmonton 

Best teams when playing away: Carolina Hurricanes and New Jersey Devils

Worst teams when playing away: Pittsburgh Penguins and Calgary Flames

## Detailed report

```{r}
# Loading packages
library(data.table)
library(ggplot2)
library(forcats)
library(stringr)

# Loading data
load('../data/2019-12-07_nhl-cleaned-data.RData')

# Setting up supporting function sfor plots
plot_theme <- theme_minimal(12) +    
  theme(axis.text.y = element_text(size = rel(1)),
        axis.ticks.y = element_blank(),
        # axis.title.x = element_text(size = rel(.75)),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(size = 0.5),
        panel.grid.minor.x = element_blank())

caption_source <- labs(caption="Note: data from regular season games\nSource: NHL.com API")

mid_line <- geom_hline(yintercept=.5)
```


### How big of an advantage do home teams have?

Overall, home teams have a significant advantage, winning ~55% of the time. This advantage has been consistent over the time period of this analysis.

```{r}
overall_win_perc_HoA_by_season <- vF_game_info[,.(h=sum(home.win),a=sum(away.win)), by = season
                             ][,.(season,home=h/(h+a),away=a/(h+a))]

ggplot(melt(overall_win_perc_HoA_by_season, id.vars='season', variable.name='HoA', value.name='win_perc')) +
  geom_point(aes(x=season,y=win_perc,color=HoA)) +
  mid_line +
  labs(title = "NHL home vs away win-percentage (2014-18)"
       ,x = "Season"
       ,y = "Win percentage"
       ,color = element_blank()) +
  caption_source +
  scale_y_continuous(limits = c(.4,.6), labels = scales::percent_format(1)) +
  plot_theme
```

Looking across teams, we see a significant variation in this performance. For example, the Tampa Bay Lightning win almost 70% of their home games and 55% of their away ones. In general, teams that perform better at home also perform better on the road. However, we do see a big difference in the how much better teams perform when they are home vs. away. The San Jose Sharks (shown in the middle) have very small difference while the Philadelphia Flyers have a massive one (almost 20%).

```{r}
team_win_perc_HoA <- vF_game_teams_stats[,.(win_perc=sum(won)/.N),by=.(team.id,HoA)]

ggplot(merge(team_win_perc_HoA, vF_teams_DT[,.(team.id,long.name)], by='team.id')) +
  geom_point(aes(x=reorder(long.name,win_perc*(HoA=='home')),y=win_perc,color=HoA)) +
  mid_line +
  labs(title = "Home vs away win-percentage by team (2014-18)"
       ,x=element_blank()
       ,y="Win percentage"
       ,color=element_blank()) +
  caption_source +
  scale_y_continuous(limits = c(.3,.7), labels = scales::percent_format(1)) +
  plot_theme +
  theme(axis.text.y = element_text(size=rel(.75))) +
  coord_flip()
```

Clearly we need to account for this variation in team-strength. As we dig below the level of wins and losses, we will also explore metrics that adjust for this.

### What does home team advantage look like for different types of events?

A simple view of what percentage of plays are recorded for the home team highlights how the home team plays better. On the defensive side, home teams have more takeaways and hits. However, they block fewer shots, which could be explained by the fact that away teams shoot less. On the offensive side, home teams have a majority of the plays across all types. 

And lastly, the other category is a little more interesting. Fewer penalties are called on the home team (in this case <50% is better for the home team). Perhaps referees are less inclined to call a penalty when the entire stadium is yelling at them for it. Additionally, home teams give away the puck much more than away teams. 

```{r}
play_type <- data.table(result.eventTypeId=c('BLOCKED_SHOT','FACEOFF','GIVEAWAY','GOAL','HIT','MISSED_SHOT','PENALTY','SHOT','TAKEAWAY'),
type=c('Defensive','Other','Other','Offensive','Defensive','Offensive','Other','Offensive','Defensive'))
event_home_perc <- dcast(vF_game_plays[,.N,by=.(HoA,result.eventTypeId)], result.eventTypeId ~ HoA)[,.(result.eventTypeId,home_perc = H/(A+H))]

ggplot(merge(event_home_perc, play_type)) +
  geom_col(aes(x=reorder(str_to_sentence(result.eventTypeId),home_perc), y=home_perc)) +
  mid_line +
  labs(title = "Percentage of plays for the home team (2014-18)"
       ,x=element_blank()
       ,y="Home plays as a % of total") +
  caption_source +
  scale_y_continuous(labels = scales::percent_format(1)) +
  facet_grid(rows=vars(type), scales = "free_y", space = "free_y") +
  plot_theme +
  coord_flip()

```

But what does this look like when accounting for team strength? The boxplots below show how teams perform at home compared to away for all play types. This metric is calculated as 
$$\frac{\text{avg # plays when home per game} - \text{avg # plays when away per game}} {\text{avg # plays when away per game}}$$
This view provides a bit more nuance to the story. Some play types (like giveaways) have a large amount of variability as compared to goals, where all teams clearly perform better at home.

```{r}
plays_summary <- vF_game_plays[,.(rate_per_game = .N / length(unique(game.id))),by=.(team.id.for, HoA,result.eventTypeId)]

plays_home_delta <- dcast(plays_summary, team.id.for + result.eventTypeId ~ HoA)[,perc_delta := (H-A)/A]

ggplot(merge(plays_home_delta,play_type,by="result.eventTypeId")) +
  geom_boxplot(aes(x=reorder(str_to_sentence(result.eventTypeId),perc_delta,median), y=perc_delta)) +
  geom_hline(yintercept=0) +
  labs(title = "Difference in performance home vs. away (2014-18)"
       ,subtitle = "Each observation represents one team for that type of play"
       ,x=element_blank()
       ,y="Home performance relative to away performance") +
  caption_source +
  scale_y_continuous(limits=c(-.5,1.25), labels = scales::percent_format(1)) +
    facet_grid(rows=vars(type), scales = "free_y", space = "free_y") +
  plot_theme +
  coord_flip()
```

### What arenas are the hardest to play at?

To understand this question, we look at each arena and how other teams perform when they play there. We then compare this performance to that team's average away performance. The three hardest places to play are Anaheim, Washington, and Montreal. Interestingly, in a recent poll of NHL players, Nashville, Winnipeg, and Pittsburgh were ranked as the toughest places to play. While these three are in the top half, they are certainly not in the top 5. On the other side of the coin, Ottawa and Edmonton are the easiest places to play. This is not surprising given that over this time period these teams did not perform well in general

```{r}
play_sel <- 'GOAL'
avg_away_performance <- plays_summary[HoA=='A' & result.eventTypeId==play_sel, .(team.id.for, avg_rate = rate_per_game)]

away_performance_by_arena <- vF_game_plays[HoA=='A' & result.eventTypeId==play_sel
                                           ,.(rate_per_game = .N / length(unique(game.id))),by=.(team.id.for, team.id.against)][order(team.id.for,team.id.against)]

away_perf_comp <- merge(away_performance_by_arena, avg_away_performance)[, delta := rate_per_game/avg_rate-1]

away_perf_by_city <- merge(away_perf_comp[,mean(delta),by=team.id.against], vF_teams_DT[,.(team.id,venue.city)], by.x='team.id.against', by.y='team.id')

ggplot(away_perf_by_city) +
  geom_point(aes(x=reorder(venue.city,V1),y=V1)) +
  geom_hline(yintercept=0) +
  labs(title = "Away team performance by city (2014-18)"
       ,subtitle = paste0("Perfomance measured using ",str_to_lower(play_sel),"s")
       ,x=element_blank()
       ,y="Performance relative to avg away performance") +
  caption_source +
  # scale_y_continuous(limits = c(.3,.7), labels = scales::percent_format(1)) +
  plot_theme +
  theme(axis.text.y = element_text(size=rel(.75))) +
  coord_flip()
```

Paragraph on arena visualizations (to be cleaned up and actual plots to be attached when finalized)

These are all pretty similar. The home team has a denser region of shots and goals right in front of goal.

Anaheim - Very dense crescent shape in front of goal for Ducks
Washington - Denser in front of goal for Capitals
Montreal -
Nashville - Area of danger slightly smaller but denser with goals for Predators
Winnipeg - Slightly denser for Jets but not by much
Pittsburgh - Penguins map is very dense in front of goal compared to visitors

The density of the home danger area is less pronounced for Edmonton but not for Ottawa (Ottawa supposedly the easiest to play away). Ottawa has a denser area in front of goal than their visitors.

### Case study: 

### What teams out-perform when on the road?

Earlier, we looked at win rates when teams are at home vs. away and noticed significant variation in the difference between home and away performance. We now look at this, using goals, while accounting for team strength. The Carolina Hurricanes, New Jersey Devils, and San Jose Sharks perform the best when away. Earlier we noticed the San Jose Sharks had a very small difference in home and away win percentage. But while the Carolina Hurricanes difference is small here, their home vs. away win percentage is much larger. (Of course, scoring goals is only half the battle in hockey!) 

And on the other side, the Pittsburgh Penguins perform the worst when away. This is interesting because they were a really strong team over this time period. But their overall home/away win percentages do reflect this.

```{r}
play_sel <- 'GOAL'

plays_away_delta <- dcast(plays_summary, team.id.for + result.eventTypeId ~ HoA)[,perc_delta := (A-H)/H]

team_away_perf <- merge(plays_away_delta,vF_teams_DT[,.(team.id,long.name)], by.x='team.id.for', by.y='team.id')

ggplot(team_away_perf[result.eventTypeId == play_sel]) +
  geom_point(aes(x=reorder(long.name,perc_delta),y=perc_delta)) +
  geom_hline(yintercept=0) +
  labs(title = "Away performance relative to home performance (2014-18)"
       ,subtitle = paste0("Perfomance measured using ",str_to_lower(play_sel),"s")
       ,x=element_blank()
       ,y="Away performance as a % of home performance") +
  caption_source +
  scale_y_continuous(labels = scales::percent_format(1)) +
  plot_theme +
  theme(axis.text.y = element_text(size=rel(.75))) +
  coord_flip()

```

Paragraph on team home v away visualizations (to be cleaned up and actual plots to be attached when finalized)

Teams that perform relatively well away
- Carolina Hurricanes: When playing away their shotmap is a bit more compact but shots close to goal less dense. Overall not too much of a difference as expected
- New Jersey Devils: Home map is more compact with a lot of shots coming from the left side. 
- San Jose Sharks: Home map has a good amount of goals coming from the left.

Teams that perform relatively worse away
- Pittsburgh Penguins: Very dense around the goal when playing at home, much more spread out when playing away
- Calgary Flames: Again pretty dense around goal when playing at home but patterns arent as clear
- Winnipeg Jets: Pattern isnt too clear

Teams that have interesting patterns
- Philadelphia Flyers: Home plot shows shots clustered just past the blue line(dont know what its called) and halfway between that blue line and goal. Suggests they have a structured way of playing at home but not away
- Tampa Bay Lightning: Home plot is very dense and compact. Still strong when playing away but not to the same extent
- Boston Bruins: Home plot shows very dense number of shots from far out to either side, not so when playing away


### Case study: 
